---
title: "asv_generation_its"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


# **Project Summary** 

The following code describes the sequence processing steps used to analyze **v4** data generated for the **washington elevation gradient** project. Data comes from DNA extracted from soils and surface sterilized roots. Samples were collected from Washington transect in **late summer 2023**. 

**Key**

**Plant Compartment**

1. **Rhizosphere Soil** == rhizo
2. **Root Endosphere** == root

**Amplicon**

1. **Bacteria** == o16S
2. **Fungi** == oITS

**Site**
1. CR1
2. CR1.5
3. CR2
4. CR3
5. CR4
6. TR1
7. TR2
8. TR2.5
9. TR3
10. TR3.5
11. TR4

**Temperature and Rainfall**

1. **Cool and wet on the east side of the Cascades, Cowlitz River Watershed** == cr
2. **Hot and dry on the west side of the Cascades, Tieton River Watershed** == tr

# **Load Packages**

First install (if needed)
```{r}
install.packages("BiocManager")
install.packages("Biostrings")
install.packages("ShortRead")
install.packages("reticulate")
install.packages("tidyverse")
BiocManager::install("dada2")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("dada2", version = "3.11")
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.20")
install.packages("devtools")
library("devtools")
devtools::install_github("benjjneb/dada2", ref="v1.16") # change the ref argument to get other versions
```


Loading the packages required to complete the sequence processing. 
```{r}
#Load Packages
library(dada2)
library(Biostrings)
library(ShortRead)
library(reticulate)
library(tidyverse)
```

# Trimming Primers

Removing primers using cutadapt. 

## Cutadapt Flags and system connection

Creating a function to generate primer flags for cutadapt. The function takes the following arguments:

**amplicon**: The function comes pre-loaded with the most common primer sets used in the lab. This includes *its2_cregger* and *v4_cregger*. The default is *NULL* on the off-chance that another primer set is used. 

**primer_fwd_list**: If amplicon is NULL then this is a list of user specified forward primer(s).

**primer_rev_list**: If amplicon is NULL then this is a list of user specified reverse primer(s). 

```{r}
# Creating the primer_strings function (if there are custom primers)
primer_strings<-function(amplicon=v4_cregger, # choose preset amplicon or write NULL
                         primer_fwd_list=NULL,primer_rev_list=NULL){ # fill in custom sequence or keep as NULL
  if(is.null(amplicon)==FALSE && !amplicon=="its2_cregger" && !amplicon=="v4_cregger"){
    stop("amplicon should be its2_cregger or v4_cregger. If using non-cregger lab primers leave the amplicon argument blank and specify primers in the primer_fwd_list and primer_rev_list")
  }
  cutadapt_primer_flags<-list()
  if(is.null(amplicon)==FALSE && amplicon=="its2_cregger"){  
  its3_forward<-c("CATCGATGAAGAACGCAG","CAACGATGAAGAACGCAG","CACCGATGAAGAACGCAG","CATCGATGAAGAACGTAG","CATCGATGAAGAACGTGG","CATCGATGAAGAACGCTG")
    its4_reverse<-c("TCCTSCGCTTATTGATATGC","TCCTCGCCTTATTGATATGC")
    its3_forward_rc<-lapply(its3_forward,dada2::rc)
    its4_reverse_rc<-lapply(its4_reverse,dada2::rc)
    for (i in 1:length(its3_forward)){
      for(x in 1:length(its4_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",its3_forward[i],"...",its4_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(its4_reverse)){
      for(x in 1:length(its3_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",its4_reverse[i],"...",its3_forward_rc[x], sep=''))
      }
    }    
  }else if(is.null(amplicon)==FALSE && amplicon=="v4_cregger"){
    v4_forward<-c("GTGCCAGCMGCCGCGGTAA","GTGCCAGCMGCWGCGGTAA","GTGCCAGCMGCCGCGGTCA","GTGKCAGCMGCCGCGGTAA")
    v4_reverse<-c("GGACTACHVGGGTWTCTAAT")
    v4_forward_rc<-lapply(v4_forward,dada2::rc)
    v4_reverse_rc<-lapply(v4_reverse,dada2::rc)
    for (i in 1:length(v4_forward)){
      for(x in 1:length(v4_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",v4_forward[i],"...",v4_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(v4_reverse)){
      for(x in 1:length(v4_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",v4_reverse[i],"...",v4_forward_rc[x], sep=''))
      }
    } 
  }else if(is.null(amplicon)==TRUE){
    primer_forward<-primer_fwd_list
    primer_reverse<-primer_rev_list
    primer_forward_rc<-lapply(primer_forward,dada2::rc)
    primer_reverse_rc<-lapply(primer_reverse,dada2::rc)
    for (i in 1:length(primer_forward)){
      for(x in 1:length(primer_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",primer_forward[i],"...",primer_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(primer_reverse)){
      for(x in 1:length(primer_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",primer_reverse[i],"...",primer_forward_rc[x], sep=''))
      }
    } 
  }
    return(cutadapt_primer_flags)
}

# Specifies the conda environment to use for cutadapt (update to conda environment location) #reticulate
use_condaenv("/Users/8qq/miniconda3/envs/cutadapt")

# Uses the system2 function to pass commands to the shell to run cutadapt from R. Checking to make sure R can find Cutadapt. 
system2("conda", args = c("run", "-n", "cutadapt", "cutadapt", "--version"))

```


# **Path Designations**

Designating the paths for the directory that contains...

1. 16S Rhizosphere sequencing data = o16S_rhizo
2. 16S Root sequencing data = o16S_root
3. 16S Soil sequencing data = o16S_soil
4. ITS Rhizosphere sequencing data = oITS_rhizo
5. ITS Root sequencing data = oITS_root
6. ITS Soil sequencing data = oITS_soil


**1. 16S Rhizosphere sequencing data**
```{r}
# Creating an object that has the pathway for the sequence directory
o16S_rhizo <- "/Users/8qq/Desktop/R/incline/16s_rhizo_raw/"

# Listing the files present in the sequence directory
list.files(o16S_rhizo)

# Creating objects for the forward and reverse reads
f_o16S_rhizo <- sort(list.files(o16S_rhizo, pattern="R1_001",full.names=TRUE))
r_o16S_rhizo <- sort(list.files(o16S_rhizo, pattern="R2_001",full.names=TRUE))

# Checking number of forward and reverse files. Should be an equal number of forward and reverse files.
length(f_o16S_rhizo) #9... what is the difference between rhizo and soil? The 16S soil dataset is mostly missing?
length(r_o16S_rhizo) #7

# Creating an object  that strips away the pathway information leaving behind only the filename.
filebasename_o16S_rhizo <- basename(f_o16S_rhizo)

basename_o16S_rhizo <- sub("-16s.*","_roots", filebasename_o16S_rhizo)

# Using gsub to substitute the - for the _ which is used in the metadata sheet. 
samplenames_o16S_rhizo <- gsub("\\-","_", basename_o16S_rhizo)

unique(samplenames_o16S_rhizo)
```

**2. 16S Root sequencing data**

```{r}
# Creating an object that has the pathway for the sequence directory
o16S_root <- "/Users/8qq/Desktop/R/incline/roots_raw/16s_roots_raw"

# Listing the files present in the sequence directory
list.files(o16S_root)

# Creating objects for the forward and reverse reads
f_o16S_root <- sort(list.files(o16S_root, pattern = "R1_001", 
                               full.names = TRUE))
r_o16S_root <- sort(list.files(o16S_root, pattern = "R2_001", 
                               full.names = TRUE))

# Checking number of forward and reverse files. Should be an equal number of forward and reverse files.
length(f_o16S_root) #33
length(r_o16S_root) #33

# Creating an object  that strips away the pathway information leaving behind only the filename.
filebasename_o16S_root <- basename(f_o16S_root)

samplenames_o16S_root <- filebasename_o16S_root %>%
  str_replace("_L001_R1_001.fastq","") %>%
  str_replace("-5-", ".5_") %>%
  str_replace_all("-", "_")

unique(samplenames_o16S_root)
```

# Initial Quality Assessment

Assessing quality of sequences using a qc plot.

```{r}
# Generating quality plots
f_o16S_root_qual_plot <- plotQualityProfile(f_o16S_root, aggregate = TRUE)
f_o16S_root_qual_plot
r_o16S_root_qual_plot <- plotQualityProfile(r_o16S_root, aggregate = TRUE)
r_o16S_root_qual_plot
```

# Initial Quality Filtering

Removing reads with ambiguous bases prior to using cutadapt to remove primers. 

```{r}
# Creating two path objects that will store the paths for the filtered forward and reverse reads. 
f_o16S_root_filtN <- file.path(o16S_rhizo, "filtN", basename(f_o16S_root))
r_o16S_root_filtN <- file.path(o16S_rhizo, "filtN", basename(r_o16S_root))

# Filtering forward and reverse reads using the filterAndTrim function. Specifying maxN = 0 will indicate that sequences with at least 1 ambiguous base will be removed from the dataset. 
filterAndTrim(f_o16S_root, f_o16S_root_filtN, r_o16S_root, 
              r_o16S_root_filtN, maxN = 0, multithread = TRUE) #long

# Creates quality profiles for the pre-filtered forward and reverse reads.
f_o16S_root_filtN_qual_plot <- plotQualityProfile(f_o16S_root_filtN, 
                                                  aggregate = TRUE) #long
f_o16S_root_filtN_qual_plot 
r_o16S_root_filtN_qual_plot <- plotQualityProfile(r_o16S_root_filtN,
                                                  aggregate = TRUE) #long
r_o16S_root_filtN_qual_plot
```

## Primer Trimming with Cutadapt

```{r}
# Specifying primer strings
v4_primer_strings <- primer_strings(amplicon = "v4_cregger")

# Creating a directory to store the forward and reverse reads after they have been trimmed. 
path_cut <- file.path(o16S_root, "cutadapt")
if(!dir.exists(path_cut)) dir.create(path_cut)

f_o16S_root_cut <- file.path(path_cut, basename(f_o16S_root))
r_o16S_root_cut <- file.path(path_cut, basename(r_o16S_root))

# Using cutadapt to remove the primers from each read. Removing any read that does not have the forward primer or the reverse primer in the correct orientation.

for (i in seq_along(f_o16S_root)){
  system2("conda", args = c("run -n cutadapt cutadapt",
                          v4_primer_strings,
                          "--discard-untrimmed", 
                          "--minimum-length 10",
                          "--report=minimal",
                          "-o", f_o16S_root_cut[i], 
                          "-p", r_o16S_root_cut[i],
                  f_o16S_root_filtN[i],
                  r_o16S_root_filtN[i]))
}

```

# Filtering the cut Sequences based on quality

```{r}
# Creating new pathways for the quality filtered forward and reverse reads. 
f_o16S_root_filt <- file.path(o16S_root, "filtered", 
                              (paste0(samplenames_o16S_root, 
                                     "_F_filt.fastq")))
r_o16S_root_filt <- file.path(o16S_root, "filtered", 
                              paste0(samplenames_o16S_root, 
                                     "_R_filt.fastq")) 

# Creating quality profiles for the primer trimmed forward and reverse reads. 
f_o16S_root_cut_qual_plot <- 
  plotQualityProfile(f_o16S_root_cut, aggregate = TRUE) #long

f_o16S_root_cut_qual_plot + 
  geom_vline(aes(xintercept = 230))

r_o16S_root_cut_qual_plot <- 
  plotQualityProfile(r_o16S_root_cut, aggregate = TRUE) #long

r_o16S_root_cut_qual_plot + 
  geom_vline(aes(xintercept = 230))

# Quality filtering reads using a maximum expected error threshold of 2 for the forward and reverse reads. 
o16S_root_filter.out <- filterAndTrim(fwd = f_o16S_root_cut, 
                                     filt = f_o16S_root_filt, 
                                     rev = r_o16S_root_cut, 
                                     filt.rev = r_o16S_root_filt, 
                                     maxEE = c(2, 2), 
                                     compress = FALSE, truncLen = 230, 
                                     multithread = TRUE) #long
 
# Converting the filter.out_v4_soil_p2 object into a data frame. 
o16S_root_filter.out <- as.data.frame(o16S_root_filter.out)

# Checking to see the effects of the quality filtering on per sample read counts. 
o16S_root_filter.out$diffs <- 
  o16S_root_filter.out$reads.in-o16S_root_filter.out$reads.out

# Using the order command to sort the dataframe by the new column (diffs) that created above. 
o16S_root_filter.out[order(o16S_root_filter.out$reads.out),]

# Removing files that were not written as a result of no reads passing filter
f_o16S_root_filt_no_missing <- subset(f_o16S_root_filt, basename(f_o16S_root_filt)%in%list.files(file.path(o16S_root, "filtered")))

r_o16S_root_filt_no_missing <- subset(r_o16S_root_filt, basename(r_o16S_root_filt)%in%list.files(file.path(o16S_root, "filtered")))

# Generating quality plots for the forward and reverse reads after quality filtering
f_o16S_root_filt_no_missing_qual_plot <- plotQualityProfile(f_o16S_root_filt_no_missing, aggregate = TRUE)
f_o16S_root_filt_no_missing_qual_plot
r_o16S_root_filt_no_missing_qual_plot <- plotQualityProfile(r_o16S_root_filt_no_missing, aggregate = TRUE)
r_o16S_root_filt_no_missing_qual_plot
```

(past this point it doesn't get put into a new directory just saved in the R environment. There will be objects to open and variables to change throughout.)

## DADA Denoising

Denoising data by first estimating sequencing error rates and using the dada algorithm. 

```{r}
# De-replicate forward and reverse reads
f_o16S_root_derep <- derepFastq(f_o16S_root_filt_no_missing, verbose = TRUE)
r_o16S_root_derep <- derepFastq(r_o16S_root_filt_no_missing, verbose = TRUE)

# Assigning the sample names to the de-replicated sequence objects
names(f_o16S_root_derep) <- subset(samplenames_o16S_root, 
                                       samplenames_o16S_root%in%sub("_F_filt.fastq", "", 
                                          names(f_o16S_root_derep)))
names(r_o16S_root_derep) <- subset(samplenames_o16S_root, 
                                       samplenames_o16S_root%in%sub("_R_filt.fastq", "", 
                                          names(r_o16S_root_derep)))

# Learning errors for forward and reverse reads. 
f_o16S_root_error <- learnErrors(f_o16S_root_derep, multithread = TRUE)
r_o16S_root_error <- learnErrors(r_o16S_root_derep, multithread = TRUE)

# Error plots for forward & reverse reads
f_o16S_root_error_plot <- plotErrors(f_o16S_root_error, nominalQ = TRUE)
f_o16S_root_error_plot
r_o16S_root_error_plot <- plotErrors(r_o16S_root_error, nominalQ = TRUE)
r_o16S_root_error_plot

# De-noise forward reads with dada2
f_o16S_root_dada <- dada(f_o16S_root_derep, err = f_o16S_root_error, 
                         multithread = TRUE, pool = TRUE) # long! Can perform in QIIME2 if it won't run here, but R gives better diversity estimates because it doesn't filter out singletons

# De-noise reverse reads with dada2
r_o16S_root_dada <- dada(r_o16S_root_derep, err = r_o16S_root_error, 
                         multithread = TRUE, pool = TRUE) # long! Can perform in QIIME2 if it won't run here, but R gives better diversity estimates because it doesn't filter out singletons

```


# Merging the de-noised reads with mergePairs

```{r}
# Merging forward and reverse reads using mergePairs.
o16S_root_merge <- mergePairs(f_o16S_root_dada, f_o16S_root_derep, r_o16S_root_dada,
                              r_o16S_root_derep, verbose = TRUE)

# Creating a sequence table
o16S_root_seqtab <- makeSequenceTable(o16S_root_merge)
# histogram to check average sequence length
hist(nchar(colnames(o16S_root_seqtab)))

# Merging ASV tables from all four runs # skipping this step for now but will revisit if I want to combine runs in the future
# merged_sequence_table<-mergeSequenceTables(seqtab_v4_roots,seqtab_v4_soil_redo_d10.d25,seqtab_v4_soil_redo_initial)

# Removing chimeras using the removeBimeraDenovo function.
o16S_root_seqtab_nochim <- removeBimeraDenovo(o16S_root_seqtab, method = "consensus", 
                                         multithread = TRUE, verbose = TRUE)
# To run on the merged seq table:
# o16S_seqtab_nochim <- removeBimeraDenovo(merged_sequence_table, method = "consensus",  multithread = TRUE, verbose = TRUE)

# Number of ASVs in the table:
ncol(o16S_root_seqtab_nochim)
# Number of samples in the table:
nrow(o16S_root_seqtab_nochim)

# Number of sequences
sum(o16S_root_seqtab_nochim)
```

# Assign taxonomy with assignTaxonomy. Update SILVA and UNITE each time
Could not run SILVA because I could not install ARB due to homebrew issues. I tried: 
bash
softwareupdate --install-rosetta
arch -x86_64 /bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install.sh)"
arch -x86_64 brew install arb-project/arb/arb

bash
softwareupdate --install-rosetta
arch -x86_64 /usr/local -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install.sh)"
arch -x86_64 brew install arb-project/arb/arb

With the following errors:
Error: Cannot install under Rosetta 2 in ARM default prefix (/opt/homebrew)!
To rerun under ARM use:
    arch -arm64 brew install ...
To install under x86_64, install Homebrew into /usr/local.
bash-3.2$ arch -arm64 brew install arb-project/arb/arb
arb: The x86_64 architecture is required for this software.
Error: arb: An unsatisfied requirement failed this build.

Instead, I used the .fa training file on the DADA2 github.
https://benjjneb.github.io/dada2/training.html

```{r}
# Assigning taxonomy using the assignTaxonomy function. 
# Update SILVA and UNITE each time to make sure we are getting the latest assignments 
# Assign sequences with SILVA:
o16S_root_taxa <- assignTaxonomy(o16S_root_seqtab_nochim, 
                          "~/Desktop/R/silva_nr99_v138.2_toSpecies_trainset.fa.gz", 
                          multithread = TRUE, minBoot = 80) # long
# See for example of path to OneDrive:
# taxa_v4 <- assignTaxonomy(seqtab.nochim_v4, "/Users/5io/Library/CloudStorage/OneDrive-OakRidgeNationalLaboratory/septoria_in_a_bag/amplicon_sequencing/soil_amplicon_sequencing/v4_data/silva_nr99_v138.1_wSpecies_train_set.fa.gz", multithread = TRUE, minBoot = 80)

# Converting the taxonomy assignments to data frame so we can filter the taxonomic assignments.
o16S_root_taxa_original <- as.data.frame(o16S_root_taxa)

# Filtering out ASVs not assigned to a phyla
o16S_root_taxa_no_na <- o16S_root_taxa_original[-(which(is.na(o16S_root_taxa_original$Phylum))),]

# Filtering out ASVs not assigned to bacteria or archaea
o16S_root_taxa_bac <- o16S_root_taxa_no_na[grep("Bacteria|Archaea", 
                                                      o16S_root_taxa_no_na$Kingdom),]

# Filtering out ASVs assigned to a chloroplast
o16S_root_taxa_nochl <- o16S_root_taxa_bac[grep("Chloroplast", o16S_root_taxa_bac$Order, 
                                                invert = TRUE),]

# Filtering out ASVs assigned to a mitochondria
o16S_root_taxa_nohost <- o16S_root_taxa_nochl[grep("Mitochondria", o16S_root_taxa_nochl$Family, 
                                                   invert=TRUE),]

```

## **Taxonomy Merging and ASV Filtering**
The last step is to filter the ASV abundance table by merging the ASV table with the taxonomy assignments. This will remove ASVs from the count table that were not assigned to a fungal phyla or were identified as nematodes or plant sequences. 

```{r}
# Transposing the ASV table so that taxonomy can be added. 
t_o16S_root_seqtab_nochim <- t(o16S_root_seqtab_nochim)

# Merging the two tables together based on row name. 
t_o16S_root_seqtab_filt <- 
  t_o16S_root_seqtab_nochim[row.names(t_o16S_root_seqtab_nochim)%in%row.names(
    o16S_root_taxa_nohost),]

colnames(t_o16S_root_seqtab_filt) <- sub("16s_S[0-9].*", "16s_roots", colnames(t_o16S_root_seqtab_nochim))

# Number of ASVs & number of sequences post-filtering
nrow(t_o16S_root_seqtab_filt)
sum(t_o16S_root_seqtab_filt)

# Merging taxonomy information into ASV table
t_o16S_root_seqtab_tax <- merge(t_o16S_root_seqtab_filt, 
                                o16S_root_taxa_nohost, by = "row.names")

#  Creating ASV labels and make these new row names.
asv_o16S_root <- as.character(c(1:nrow(t_o16S_root_seqtab_tax)))
asv_o16S_root <- paste0("asv_o16S_root_", labels(asv_o16S_root))

row.names(t_o16S_root_seqtab_tax) <- NULL
row.names(t_o16S_root_seqtab_tax) <- asv_o16S_root

write.table(t_o16S_root_seqtab_tax, "~/Desktop/R/incline/roots_raw/16s_roots_raw/o16S_root_otutab.csv", 
            row.names = TRUE, col.names = TRUE, sep = ',')

# couldn't figure out what this was for
# sort(colSums(select_if(t_o16S_root_seqtab_tax, is.numeric)))

```


############################ ITS Root Analysis #################################

**4. ITS Root Data**

## Cutadapt Flags and system connection

Creating a function to generate primer flags for cutadapt. The function takes the following arguments:

**amplicon**: The function comes pre-loaded with the most common primer sets used in the lab. This includes *its2_cregger* and *v4_cregger*. The default is *NULL* on the off-chance that another primer set is used. 

**primer_fwd_list**: If amplicon is NULL then this is a list of user specified forward primer(s).

**primer_rev_list**: If amplicon is NULL then this is a list of user specified reverse primer(s). 

```{r}
# Creating the primer_strings function (if there are custom primers)
primer_strings<-function(amplicon=its2_cregger, # choose preset amplicon or right NULL
                         primer_fwd_list=NULL,primer_rev_list=NULL){ # fill in custom sequence or keep as NULL
  if(is.null(amplicon)==FALSE && !amplicon=="its2_cregger" && !amplicon=="v4_cregger"){
    stop("amplicon should be its2_cregger or v4_cregger. If using non-cregger lab primers leave the amplicon argument blank and specify primers in the primer_fwd_list and primer_rev_list")
  }
  cutadapt_primer_flags<-list()
  if(is.null(amplicon)==FALSE && amplicon=="its2_cregger"){  
  its3_forward<-c("CATCGATGAAGAACGCAG","CAACGATGAAGAACGCAG","CACCGATGAAGAACGCAG",
                  "CATCGATGAAGAACGTAG","CATCGATGAAGAACGTGG","CATCGATGAAGAACGCTG")
    its4_reverse<-c("TCCTSCGCTTATTGATATGC","TCCTCGCCTTATTGATATGC")
    its3_forward_rc<-lapply(its3_forward,dada2::rc)
    its4_reverse_rc<-lapply(its4_reverse,dada2::rc)
    for (i in 1:length(its3_forward)){
      for(x in 1:length(its4_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",its3_forward[i],"...",its4_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(its4_reverse)){
      for(x in 1:length(its3_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",its4_reverse[i],"...",its3_forward_rc[x], sep=''))
      }
    }    
  }else if(is.null(amplicon)==FALSE && amplicon=="v4_cregger"){
    v4_forward<-c("GTGCCAGCMGCCGCGGTAA","GTGCCAGCMGCWGCGGTAA",
                  "GTGCCAGCMGCCGCGGTCA","GTGKCAGCMGCCGCGGTAA")
    v4_reverse<-c("GGACTACHVGGGTWTCTAAT")
    v4_forward_rc<-lapply(v4_forward,dada2::rc)
    v4_reverse_rc<-lapply(v4_reverse,dada2::rc)
    for (i in 1:length(v4_forward)){
      for(x in 1:length(v4_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",v4_forward[i],"...",v4_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(v4_reverse)){
      for(x in 1:length(v4_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",v4_reverse[i],"...",v4_forward_rc[x], sep=''))
      }
    } 
  }else if(is.null(amplicon)==TRUE){
    primer_forward<-primer_fwd_list
    primer_reverse<-primer_rev_list
    primer_forward_rc<-lapply(primer_forward,dada2::rc)
    primer_reverse_rc<-lapply(primer_reverse,dada2::rc)
    for (i in 1:length(primer_forward)){
      for(x in 1:length(primer_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",primer_forward[i],"...",primer_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(primer_reverse)){
      for(x in 1:length(primer_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",primer_reverse[i],"...",primer_forward_rc[x], sep=''))
      }
    } 
  }
    return(cutadapt_primer_flags)
}

# Specifies the conda environment to use for cutadapt (update to conda environment location) #reticulate
use_condaenv("/Users/8qq/miniconda3/envs/cutadapt_env")

# Uses the system2 function to pass commands to the shell to run cutadapt from R. Checking to make sure R can find Cutadapt. 
system2("conda", args = c("run", "-n", "cutadapt_env", "cutadapt", "--version"))

```

# Set path to data

```{r}
# Creating an object that has the pathway for the sequence directory
oITS_root <- "/Users/8qq/Desktop/R/incline/roots_raw/its_roots_raw"

# Listing the files present in the sequence directory
list.files(oITS_root)

# Creating objects for the forward and reverse reads
f_oITS_root <- sort(list.files(oITS_root, pattern = "R1_001", 
                               full.names = TRUE))
r_oITS_root <- sort(list.files(oITS_root, pattern = "R2_001", 
                               full.names = TRUE))

# Checking number of forward and reverse files. Should be an equal number of forward and reverse files.
length(f_oITS_root) #33
length(r_oITS_root) #33

# Creating an object  that strips away the pathway information leaving behind only the filename.
filebasename_oITS_root <- basename(f_oITS_root)

samplenames_oITS_root <- filebasename_oITS_root %>%
  str_replace("_L001_R1_001.fastq","") %>%
  str_replace("-5-", ".5_") %>%
  str_replace_all("-", "_")

unique(samplenames_oITS_root)
```

# Initial Quality Assessment

Assessing quality of sequences using a qc plot.

```{r}
# Generating quality plots
f_oITS_root_qual_plot <- plotQualityProfile(f_oITS_root, aggregate = TRUE)
f_oITS_root_qual_plot
r_oITS_root_qual_plot <- plotQualityProfile(r_oITS_root, aggregate = TRUE)
r_oITS_root_qual_plot
```

# Initial Quality Filtering

Removing reads with ambiguous bases prior to using cutadapt to remove primers. 

```{r}
# Creating two path objects that will store the paths for the filtered forward and reverse reads. 
f_oITS_root_filtN <- file.path(oITS_root, "filtN", basename(f_oITS_root))
r_oITS_root_filtN <- file.path(oITS_root, "filtN", basename(r_oITS_root))

# Filtering forward and reverse reads using the filterAndTrim function. Specifying maxN = 0 will indicate that sequences with at least 1 ambiguous base will be removed from the dataset. 
filterAndTrim(f_oITS_root, f_oITS_root_filtN, r_oITS_root, 
              r_oITS_root_filtN, maxN = 0, multithread = TRUE) #long

# Creates quality profiles for the pre-filtered forward and reverse reads.
f_oITS_root_filtN_qual_plot <- plotQualityProfile(f_oITS_root_filtN, 
                                                  aggregate = TRUE) #long
f_oITS_root_filtN_qual_plot 
r_oITS_root_filtN_qual_plot <- plotQualityProfile(r_oITS_root_filtN,
                                                  aggregate = TRUE) #long
r_oITS_root_filtN_qual_plot # ~4 million reads
```

## Primer Trimming with Cutadapt

```{r}
# Specifying primer strings
its2_primer_strings <- primer_strings(amplicon = "its2_cregger")

# Creating a directory to store the forward and reverse reads after they have been trimmed. 
path_cut <- file.path(oITS_root, "cutadapt")
if(!dir.exists(path_cut)) dir.create(path_cut)

f_oITS_root_cut <- file.path(path_cut, basename(f_oITS_root))
r_oITS_root_cut <- file.path(path_cut, basename(r_oITS_root))

# Using cutadapt to remove the primers from each read. Removing any read that does not have the forward primer or the reverse primer in the correct orientation.

for (i in seq_along(f_oITS_root)){
  system2("conda", args = c("run -n cutadapt_env cutadapt",
                          its2_primer_strings,
                          "--discard-untrimmed", 
                          "--minimum-length 10",
                          "--report=minimal",
                          "-o", f_oITS_root_cut[i], 
                          "-p", r_oITS_root_cut[i],
                  f_oITS_root_filtN[i],
                  r_oITS_root_filtN[i]))
}

```

# Filtering the cut Sequences based on quality

```{r}
# Creating new pathways for the quality filtered forward and reverse reads. 
f_oITS_root_filt <- file.path(oITS_root, "filtered", 
                              (paste0(samplenames_oITS_root, 
                                     "_F_filt.fastq")))
r_oITS_root_filt <- file.path(oITS_root, "filtered", 
                              paste0(samplenames_oITS_root, 
                                     "_R_filt.fastq")) 

# Creating quality profiles for the primer trimmed forward and reverse reads. 
f_oITS_root_cut_qual_plot <- 
  plotQualityProfile(f_oITS_root_cut, aggregate = TRUE) #long

f_oITS_root_cut_qual_plot + 
  geom_vline(aes(xintercept = 230))
# down from ~4 million to ~.14 million reads

r_oITS_root_cut_qual_plot <- 
  plotQualityProfile(r_oITS_root_cut, aggregate = TRUE) #long

r_oITS_root_cut_qual_plot + 
  geom_vline(aes(xintercept = 230))

# Quality filtering reads using a maximum expected error threshold of 2 for the forward and reverse reads. 
oITS_root_filter.out <- filterAndTrim(fwd = f_oITS_root_cut, 
                                     filt = f_oITS_root_filt, 
                                     rev = r_oITS_root_cut, 
                                     filt.rev = r_oITS_root_filt, 
                                     maxEE = c(2, 2), 
                                     compress = FALSE, truncLen = 230, # Trying here because ITS is small
                                     multithread = TRUE) #long
 
# Converting the filter.out_v4_soil_p2 object into a data frame. 
oITS_root_filter.out <- as.data.frame(oITS_root_filter.out)

# Checking to see the effects of the quality filtering on per sample read counts. 
oITS_root_filter.out$diffs <- 
  oITS_root_filter.out$reads.in-oITS_root_filter.out$reads.out

# Using the order command to sort the dataframe by the new column (diffs) that created above. 
oITS_root_filter.out[order(oITS_root_filter.out$reads.out),] 

# Removing files that were not written as a result of no reads passing filter
f_oITS_root_filt_no_missing <- subset(f_oITS_root_filt, basename(f_oITS_root_filt)%in%list.files(file.path(oITS_root, "filtered")))

r_oITS_root_filt_no_missing <- subset(r_oITS_root_filt, basename(r_oITS_root_filt)%in%list.files(file.path(oITS_root, "filtered")))

# Generating quality plots for the forward and reverse reads after quality filtering
f_oITS_root_filt_no_missing_qual_plot <- plotQualityProfile(f_oITS_root_filt_no_missing, 
                                                            aggregate = TRUE)
f_oITS_root_filt_no_missing_qual_plot
r_oITS_root_filt_no_missing_qual_plot <- plotQualityProfile(r_oITS_root_filt_no_missing, 
                                                            aggregate = TRUE)
r_oITS_root_filt_no_missing_qual_plot
```

(past this point it doesn't get put into a new directory just saved in the R environment. There will be objects to open and variables to change throughout.)

## DADA Denoising

Denoising data by first estimating sequencing error rates and using the dada algorithm. 

```{r}
# De-replicate forward and reverse reads
f_oITS_root_derep <- derepFastq(f_oITS_root_filt_no_missing, verbose = TRUE)
r_oITS_root_derep <- derepFastq(r_oITS_root_filt_no_missing, verbose = TRUE)

# Assigning the sample names to the de-replicated sequence objects
names(f_oITS_root_derep) <- subset(samplenames_oITS_root, 
                                       samplenames_oITS_root%in%sub("_F_filt.fastq", "", 
                                          names(f_oITS_root_derep)))
names(r_oITS_root_derep) <- subset(samplenames_oITS_root, 
                                       samplenames_oITS_root%in%sub("_R_filt.fastq", "", 
                                          names(r_oITS_root_derep)))

# Learning errors for forward and reverse reads. 
f_oITS_root_error <- learnErrors(f_oITS_root_derep, multithread = TRUE)
r_oITS_root_error <- learnErrors(r_oITS_root_derep, multithread = TRUE)

# Error plots for forward & reverse reads
f_oITS_root_error_plot <- plotErrors(f_oITS_root_error, nominalQ = TRUE)
f_oITS_root_error_plot
r_oITS_root_error_plot <- plotErrors(r_oITS_root_error, nominalQ = TRUE)
r_oITS_root_error_plot

# De-noise forward reads with dada2
f_oITS_root_dada <- dada(f_oITS_root_derep, err = f_oITS_root_error, 
                         multithread = TRUE, pool = TRUE) # long! Can perform in QIIME2 if it won't run here, but R gives better diversity estimates because it doesn't filter out singletons

# De-noise reverse reads with dada2
r_oITS_root_dada <- dada(r_oITS_root_derep, err = r_oITS_root_error, 
                         multithread = TRUE, pool = TRUE) # long! Can perform in QIIME2 if it won't run here, but R gives better diversity estimates because it doesn't filter out singletons

```


# Merging the de-noised reads with mergePairs

```{r}
# Merging forward and reverse reads using mergePairs.
oITS_root_merge <- mergePairs(f_oITS_root_dada, f_oITS_root_derep, r_oITS_root_dada,
                              r_oITS_root_derep, verbose = TRUE)

# Creating a sequence table
oITS_root_seqtab <- makeSequenceTable(oITS_root_merge)
# histogram to check average sequence length
hist(nchar(colnames(oITS_root_seqtab)))
# what is the expected base pair length of the ITS sequence? This is variable, 250-450, 350 average and peak

# Merging ASV tables from all four runs # skipping this step for now but will revisit if I want to combine runs in the future
# merged_sequence_table<-mergeSequenceTables(seqtab_v4_roots,seqtab_v4_soil_redo_d10.d25,seqtab_v4_soil_redo_initial)

# Removing chimeras using the removeBimeraDenovo function.
oITS_root_seqtab_nochim <- removeBimeraDenovo(oITS_root_seqtab, method = "consensus", 
                                         multithread = TRUE, verbose = TRUE)
# To run on the merged seq table:
# oITS_seqtab_nochim <- removeBimeraDenovo(merged_sequence_table, method = "consensus",  multithread = TRUE, verbose = TRUE)

# Number of ASVs in the table:
ncol(oITS_root_seqtab_nochim)
# Number of samples in the table:
nrow(oITS_root_seqtab_nochim)

# Number of sequences
sum(oITS_root_seqtab_nochim)
```


# Assign Taxonomy with UNITE (re-download each time to make sure it is the most recent iteration)
You can find the latest databases to align with here: https://benjjneb.github.io/dada2/training.html

```{r}
# Assigning taxonomy using the assignTaxonomy function. 
# Update SILVA and UNITE each time to make sure we are getting the latest assignments 
# Assign sequences with SILVA:
# oITS_root_taxa <- assignTaxonomy(oITS_root_seqtab_nochim,  "~/Desktop/R/sh_general_release_all_04.04.2024/sh_general_release_dynamic_all_04.04.2024.fasta", 
                         # multithread = TRUE, minBoot = 80) # long
# The one above took too long so I'm going to try with the other file:
#oITS_root_taxa <- assignTaxonomy(oITS_root_seqtab_nochim,  "~/Desktop/R/sh_general_release_all_04.04.2024/sh_general_release_dynamic_all_04.04.2024_dev.fasta", 
                         # multithread = TRUE, minBoot = 80) # long
# This one above also took too long so I'll try with the fungal only:
oITS_root_taxa <- assignTaxonomy(oITS_root_seqtab_nochim,  "~/Desktop/R/sh_general_release_04.04.2024/sh_general_release_dynamic_04.04.2024_dev.fasta", 
                          multithread = TRUE, minBoot = 80) # empty :O 
# See for example of path to OneDrive:
# taxa_v4 <- assignTaxonomy(seqtab.nochim_v4, "/Users/5io/Library/CloudStorage/OneDrive-OakRidgeNationalLaboratory/septoria_in_a_bag/amplicon_sequencing/soil_amplicon_sequencing/v4_data/silva_nr99_v138.1_wSpecies_train_set.fa.gz", multithread = TRUE, minBoot = 80)

# Converting the taxonomy assignments to data frame so we can filter the taxonomic assignments.
oITS_root_taxa_original <- as.data.frame(oITS_root_taxa)

# Filtering out ASVs not assigned to a phyla
oITS_root_taxa_no_na <- oITS_root_taxa_original[-(which(is.na(oITS_root_taxa_original$Phylum))),]

# Filtering not necessary for ITS but adding the objects just in case they come up
oITS_root_taxa_bac <- oITS_root_taxa_no_na
oITS_root_taxa_nochl <- oITS_root_taxa_no_na
oITS_root_taxa_nohost <- oITS_root_taxa_no_na

```

## Taxonomy Merging and ASV Filtering

The last step is to filter the ASV abundance table by merging the ASV table with the taxonomy assignments. This will remove ASVs from the count table that were not assigned to a fungal phyla or were identified as nematodes or plant sequences. 

```{r}
# Transposing the ASV table so that taxonomy can be added. 
t_oITS_root_seqtab_nochim <- t(oITS_root_seqtab_nochim)

# Merging the two tables together based on row name. 
t_oITS_root_seqtab_filt <- 
  t_oITS_root_seqtab_nochim[row.names(t_oITS_root_seqtab_nochim)%in%row.names(
    oITS_root_taxa_nohost),]

colnames(t_oITS_root_seqtab_filt) <- sub("ITS_S[0-9].*", "ITS_roots", colnames(t_oITS_root_seqtab_nochim))

# Number of ASVs & number of sequences post-filtering
nrow(t_oITS_root_seqtab_filt)
sum(t_oITS_root_seqtab_filt)

# Merging taxonomy information into ASV table
t_oITS_root_seqtab_tax <- merge(t_oITS_root_seqtab_filt, 
                                oITS_root_taxa_nohost, by = "row.names")

#  Creating ASV labels and make these new row names.
asv_oITS_root <- as.character(c(1:nrow(t_oITS_root_seqtab_tax)))
asv_oITS_root <- paste0("asv_oITS_root_", labels(asv_oITS_root))

row.names(t_oITS_root_seqtab_tax) <- NULL
row.names(t_oITS_root_seqtab_tax) <- asv_oITS_root

write.table(t_oITS_root_seqtab_tax, "~/Desktop/R/incline/roots_raw/its_roots_raw/oITS_root_otutab.csv", 
            row.names = TRUE, col.names = TRUE, sep = ',')

# couldn't figure out what this was for
# sort(colSums(select_if(t_oITS_root_seqtab_tax, is.numeric)))

```




############################ 16S Rhizo and soil Analysis #######################

# Trimming Primers

Removing primers using cutadapt. 

## Cutadapt Flags and system connection

Creating a function to generate primer flags for cutadapt. The function takes the following arguments:

**amplicon**: The function comes pre-loaded with the most common primer sets used in the lab. This includes *its2_cregger* and *v4_cregger*. The default is *NULL* on the off-chance that another primer set is used. 

**primer_fwd_list**: If amplicon is NULL then this is a list of user specified forward primer(s).

**primer_rev_list**: If amplicon is NULL then this is a list of user specified reverse primer(s). 

```{r}
# Creating the primer_strings function (if there are custom primers)
primer_strings<-function(amplicon=v4_cregger, # choose preset amplicon or right NULL
                         primer_fwd_list=NULL,primer_rev_list=NULL){ # fill in custom sequence or keep as NULL
  if(is.null(amplicon)==FALSE && !amplicon=="its2_cregger" && !amplicon=="v4_cregger"){
    stop("amplicon should be its2_cregger or v4_cregger. If using non-cregger lab primers leave the amplicon argument blank and specify primers in the primer_fwd_list and primer_rev_list")
  }
  cutadapt_primer_flags<-list()
  if(is.null(amplicon)==FALSE && amplicon=="its2_cregger"){  
  its3_forward<-c("CATCGATGAAGAACGCAG","CAACGATGAAGAACGCAG","CACCGATGAAGAACGCAG","CATCGATGAAGAACGTAG","CATCGATGAAGAACGTGG","CATCGATGAAGAACGCTG")
    its4_reverse<-c("TCCTSCGCTTATTGATATGC","TCCTCGCCTTATTGATATGC")
    its3_forward_rc<-lapply(its3_forward,dada2::rc)
    its4_reverse_rc<-lapply(its4_reverse,dada2::rc)
    for (i in 1:length(its3_forward)){
      for(x in 1:length(its4_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",its3_forward[i],"...",its4_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(its4_reverse)){
      for(x in 1:length(its3_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",its4_reverse[i],"...",its3_forward_rc[x], sep=''))
      }
    }    
  }else if(is.null(amplicon)==FALSE && amplicon=="v4_cregger"){
    v4_forward<-c("GTGCCAGCMGCCGCGGTAA","GTGCCAGCMGCWGCGGTAA","GTGCCAGCMGCCGCGGTCA","GTGKCAGCMGCCGCGGTAA")
    v4_reverse<-c("GGACTACHVGGGTWTCTAAT")
    v4_forward_rc<-lapply(v4_forward,dada2::rc)
    v4_reverse_rc<-lapply(v4_reverse,dada2::rc)
    for (i in 1:length(v4_forward)){
      for(x in 1:length(v4_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",v4_forward[i],"...",v4_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(v4_reverse)){
      for(x in 1:length(v4_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",v4_reverse[i],"...",v4_forward_rc[x], sep=''))
      }
    } 
  }else if(is.null(amplicon)==TRUE){
    primer_forward<-primer_fwd_list
    primer_reverse<-primer_rev_list
    primer_forward_rc<-lapply(primer_forward,dada2::rc)
    primer_reverse_rc<-lapply(primer_reverse,dada2::rc)
    for (i in 1:length(primer_forward)){
      for(x in 1:length(primer_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",primer_forward[i],"...",primer_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(primer_reverse)){
      for(x in 1:length(primer_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",primer_reverse[i],"...",primer_forward_rc[x], sep=''))
      }
    } 
  }
    return(cutadapt_primer_flags)
}

# Specifies the conda environment to use for cutadapt (update to conda environment location) #reticulate
use_condaenv("/Users/8qq/miniconda3/envs/cutadapt")

# Uses the system2 function to pass commands to the shell to run cutadapt from R. Checking to make sure R can find Cutadapt. 
system2("conda", args = c("run", "-n", "cutadapt", "cutadapt", "--version"))

```



**1. 16S Rhizo and soil sequencing data**
We will save it as 016S_rhizo

```{r}
# Creating an object that has the pathway for the sequence directory
o16S_rhizo <- "/Users/8qq/Desktop/R/incline/16S_soil_rhizo_rerun_labkey/raw"

# Listing the files present in the sequence directory
list.files(o16S_rhizo)

# Creating objects for the forward and reverse reads
f_o16S_rhizo <- sort(list.files(o16S_rhizo, pattern = "R1_001", 
                               full.names = TRUE))
r_o16S_rhizo <- sort(list.files(o16S_rhizo, pattern = "R2_001", 
                               full.names = TRUE))

# Checking number of forward and reverse files. Should be an equal number of forward and reverse files.
length(f_o16S_rhizo) #34
length(r_o16S_rhizo) #34

# Creating an object  that strips away the pathway information leaving behind only the filename.
filebasename_o16S_rhizo <- basename(f_o16S_rhizo)

samplenames_o16S_rhizo <- filebasename_o16S_rhizo %>%
  str_replace("_L001_R1_001.fastq","") %>%
  str_replace("-5-", ".5_") %>%
  str_replace_all("-", "_")

unique(samplenames_o16S_rhizo)
```

# Initial Quality Assessment

Assessing quality of sequences using a qc plot.

```{r}
# Generating quality plots
f_o16S_rhizo_qual_plot <- plotQualityProfile(f_o16S_rhizo, aggregate = TRUE)
f_o16S_rhizo_qual_plot
r_o16S_rhizo_qual_plot <- plotQualityProfile(r_o16S_rhizo, aggregate = TRUE)
r_o16S_rhizo_qual_plot
```

# Initial Quality Filtering

Removing reads with ambiguous bases prior to using cutadapt to remove primers. 

```{r}
# Creating two path objects that will store the paths for the filtered forward and reverse reads. 
f_o16S_rhizo_filtN <- file.path(o16S_rhizo, "filtN", basename(f_o16S_rhizo))
r_o16S_rhizo_filtN <- file.path(o16S_rhizo, "filtN", basename(r_o16S_rhizo))

# Filtering forward and reverse reads using the filterAndTrim function. Specifying maxN = 0 will indicate that sequences with at least 1 ambiguous base will be removed from the dataset. 
filterAndTrim(f_o16S_rhizo, f_o16S_rhizo_filtN, r_o16S_rhizo, 
              r_o16S_rhizo_filtN, maxN = 0, multithread = TRUE) #long

# Creates quality profiles for the pre-filtered forward and reverse reads.
f_o16S_rhizo_filtN_qual_plot <- plotQualityProfile(f_o16S_rhizo_filtN, 
                                                  aggregate = TRUE) #long
f_o16S_rhizo_filtN_qual_plot
r_o16S_rhizo_filtN_qual_plot <- plotQualityProfile(r_o16S_rhizo_filtN,
                                                  aggregate = TRUE) #long
r_o16S_rhizo_filtN_qual_plot
```

## Primer Trimming with Cutadapt

```{r}
# Specifying primer strings
v4_primer_strings <- primer_strings(amplicon = "v4_cregger")

# Creating a directory to store the forward and reverse reads after they have been trimmed. 
path_cut <- file.path(o16S_rhizo, "cutadapt")
if(!dir.exists(path_cut)) dir.create(path_cut)

f_o16S_rhizo_cut <- file.path(path_cut, basename(f_o16S_rhizo))
r_o16S_rhizo_cut <- file.path(path_cut, basename(r_o16S_rhizo))

# Using cutadapt to remove the primers from each read. Removing any read that does not have the forward primer or the reverse primer in the correct orientation.

for (i in seq_along(f_o16S_rhizo)){
  system2("conda", args = c("run -n cutadapt cutadapt",
                          v4_primer_strings,
                          "--discard-untrimmed", 
                          "--minimum-length 10",
                          "--report=minimal",
                          "-o", f_o16S_rhizo_cut[i], 
                          "-p", r_o16S_rhizo_cut[i],
                  f_o16S_rhizo_filtN[i],
                  r_o16S_rhizo_filtN[i]))
}

```

# Filtering the cut Sequences based on quality

```{r}
# Creating new pathways for the quality filtered forward and reverse reads. 
f_o16S_rhizo_filt <- file.path(o16S_rhizo, "filtered", 
                              (paste0(samplenames_o16S_rhizo, 
                                     "_F_filt.fastq")))
r_o16S_rhizo_filt <- file.path(o16S_rhizo, "filtered", 
                              paste0(samplenames_o16S_rhizo, 
                                     "_R_filt.fastq")) 

# Creating quality profiles for the primer trimmed forward and reverse reads. 
f_o16S_rhizo_cut_qual_plot <- 
  plotQualityProfile(f_o16S_rhizo_cut, aggregate = TRUE) #long

f_o16S_rhizo_cut_qual_plot + 
  geom_vline(aes(xintercept = 230))

r_o16S_rhizo_cut_qual_plot <- 
  plotQualityProfile(r_o16S_rhizo_cut, aggregate = TRUE) #long

r_o16S_rhizo_cut_qual_plot + 
  geom_vline(aes(xintercept = 230))

# Quality filtering reads using a maximum expected error threshold of 2 for the forward and reverse reads. 
o16S_rhizo_filter.out <- filterAndTrim(fwd = f_o16S_rhizo_cut, 
                                     filt = f_o16S_rhizo_filt, 
                                     rev = r_o16S_rhizo_cut, 
                                     filt.rev = r_o16S_rhizo_filt, 
                                     maxEE = c(2, 2), 
                                     compress = FALSE, truncLen = 230, 
                                     multithread = TRUE) #long
 
# Converting the filter.out_v4_soil_p2 object into a data frame. 
o16S_rhizo_filter.out <- as.data.frame(o16S_rhizo_filter.out)

# Checking to see the effects of the quality filtering on per sample read counts. 
o16S_rhizo_filter.out$diffs <- 
  o16S_rhizo_filter.out$reads.in-o16S_rhizo_filter.out$reads.out

# Using the order command to sort the dataframe by the new column (diffs) that created above. 
o16S_rhizo_filter.out[order(o16S_rhizo_filter.out$reads.out),]

# Removing files that were not written as a result of no reads passing filter
f_o16S_rhizo_filt_no_missing <- subset(f_o16S_rhizo_filt, basename(f_o16S_rhizo_filt)%in%list.files(file.path(o16S_rhizo, "filtered")))

r_o16S_rhizo_filt_no_missing <- subset(r_o16S_rhizo_filt, basename(r_o16S_rhizo_filt)%in%list.files(file.path(o16S_rhizo, "filtered")))

# Generating quality plots for the forward and reverse reads after quality filtering
f_o16S_rhizo_filt_no_missing_qual_plot <- plotQualityProfile(f_o16S_rhizo_filt_no_missing, aggregate = TRUE)
f_o16S_rhizo_filt_no_missing_qual_plot
r_o16S_rhizo_filt_no_missing_qual_plot <- plotQualityProfile(r_o16S_rhizo_filt_no_missing, aggregate = TRUE)
r_o16S_rhizo_filt_no_missing_qual_plot
```

(past this point it doesn't get put into a new directory just saved in the R environment. There will be objects to open and variables to change throughout.)

## DADA Denoising

Denoising data by first estimating sequencing error rates and using the dada algorithm. 

```{r}
# De-replicate forward and reverse reads
f_o16S_rhizo_derep <- derepFastq(f_o16S_rhizo_filt_no_missing, verbose = TRUE)
r_o16S_rhizo_derep <- derepFastq(r_o16S_rhizo_filt_no_missing, verbose = TRUE)

# Assigning the sample names to the de-replicated sequence objects
names(f_o16S_rhizo_derep) <- subset(samplenames_o16S_rhizo, 
                                       samplenames_o16S_rhizo%in%sub("_F_filt.fastq", "", 
                                          names(f_o16S_rhizo_derep)))
names(r_o16S_rhizo_derep) <- subset(samplenames_o16S_rhizo, 
                                       samplenames_o16S_rhizo%in%sub("_R_filt.fastq", "", 
                                          names(r_o16S_rhizo_derep)))

# Learning errors for forward and reverse reads. 
f_o16S_rhizo_error <- learnErrors(f_o16S_rhizo_derep, multithread = TRUE)
r_o16S_rhizo_error <- learnErrors(r_o16S_rhizo_derep, multithread = TRUE)

# Error plots for forward & reverse reads
f_o16S_rhizo_error_plot <- plotErrors(f_o16S_rhizo_error, nominalQ = TRUE)
f_o16S_rhizo_error_plot
r_o16S_rhizo_error_plot <- plotErrors(r_o16S_rhizo_error, nominalQ = TRUE)
r_o16S_rhizo_error_plot

# [bookmark - start here]
# De-noise forward reads with dada2
f_o16S_rhizo_dada <- dada(f_o16S_rhizo_derep, err = f_o16S_rhizo_error, 
                         multithread = TRUE, pool = TRUE) # long! Can perform in QIIME2 if it won't run here, but R gives better diversity estimates because it doesn't filter out singletons

# De-noise reverse reads with dada2
r_o16S_rhizo_dada <- dada(r_o16S_rhizo_derep, err = r_o16S_rhizo_error, 
                         multithread = TRUE, pool = TRUE) # long! Can perform in QIIME2 if it won't run here, but R gives better diversity estimates because it doesn't filter out singletons

```


# Merging the de-noised reads with mergePairs

```{r}
# Merging forward and reverse reads using mergePairs.
o16S_rhizo_merge <- mergePairs(f_o16S_rhizo_dada, f_o16S_rhizo_derep, r_o16S_rhizo_dada,
                              r_o16S_rhizo_derep, verbose = TRUE)

# Creating a sequence table
o16S_rhizo_seqtab <- makeSequenceTable(o16S_rhizo_merge)
# histogram to check average sequence length
hist(nchar(colnames(o16S_rhizo_seqtab)))

# writing to the data folder
write.table(o16S_rhizo_seqtab, "~/Desktop/R/incline/16S_soil_rhizo_rerun_labkey/seqtab.txt", sep = "\t")

# reading back in to see if it is the same
o16S_rhizo_rerun_seqtab <- read.table("~/Desktop/R/incline/16S_soil_rhizo_rerun_labkey/seqtab.txt", header = TRUE, sep = '\t', na.strings = ".") #long
# check dimensions

dim(o16S_rhizo_rerun_seqtab)
# load root environment

o16S_rhizo_rerun_seqtab <- read.table("~/Desktop/R/incline/16S_soil_rhizo_rerun_labkey/seqtab.txt", header = TRUE, sep = '\t', na.strings = ".") #long!!
# check dimensions
dim(o16S_rhizo_rerun_seqtab)
class(o16S_rhizo_rerun_seqtab)
# make it a matrix
o16S_rhizo_rerun_seqtab_matrix <- as.matrix(o16S_rhizo_rerun_seqtab)

# Merging ASV tables from all four runs # skipping this step for now but will revisit if I want to combine runs in the future
o16S_merged_seqtab <- mergeSequenceTables(o16S_rhizo_rerun_seqtab_matrix, o16S_root_seqtab) #check root table

# check dimensions of original and combined table (should have added rows and combined some columns)
dim(o16S_rhizo_rerun_seqtab_matrix)
dim(o16S_root_seqtab)
dim(o16S_merged_seqtab)

# Removing chimeras using the removeBimeraDenovo function.
# o16S_rhizo_seqtab_nochim <- removeBimeraDenovo(o16S_rhizo_seqtab, method = "consensus", 
#                                         multithread = TRUE, verbose = TRUE)
# To run on the merged seq table:
o16S_seqtab_nochim <- removeBimeraDenovo(o16S_merged_seqtab, method = "consensus",  multithread = TRUE, verbose = TRUE)

# Number of ASVs in the table:
ncol(o16S_merged_seqtab)
# Number of samples in the table:
nrow(o16S_seqtab_nochim)

# Number of sequences
sum(o16S_merged_seqtab)
# Number of sequences
sum(o16S_seqtab_nochim)
```

# Assign taxonomy with assignTaxonomy. Update SILVA and UNITE each time
Could not run SILVA because I could not install ARB due to homebrew issues. I tried: 
bash
softwareupdate --install-rosetta
arch -x86_64 /bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install.sh)"
arch -x86_64 brew install arb-project/arb/arb

bash
softwareupdate --install-rosetta
arch -x86_64 /usr/local -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install.sh)"
arch -x86_64 brew install arb-project/arb/arb

With the following errors:
Error: Cannot install under Rosetta 2 in ARM default prefix (/opt/homebrew)!
To rerun under ARM use:
    arch -arm64 brew install ...
To install under x86_64, install Homebrew into /usr/local.
bash-3.2$ arch -arm64 brew install arb-project/arb/arb
arb: The x86_64 architecture is required for this software.
Error: arb: An unsatisfied requirement failed this build.

Instead, I used the .fa training file on the DADA2 github.
https://benjjneb.github.io/dada2/training.html

```{r}
# Assigning taxonomy using the assignTaxonomy function. 
# Update SILVA and UNITE each time to make sure we are getting the latest assignments 
# Assign sequences with SILVA:
o16S_taxa <- assignTaxonomy(o16S_seqtab_nochim, 
                          "~/Desktop/R/silva_nr99_v138.2_toSpecies_trainset.fa.gz", 
                          multithread = TRUE, minBoot = 80) # long
# See for example of path to OneDrive:
# taxa_v4 <- assignTaxonomy(seqtab.nochim_v4, "/Users/5io/Library/CloudStorage/OneDrive-OakRidgeNationalLaboratory/septoria_in_a_bag/amplicon_sequencing/soil_amplicon_sequencing/v4_data/silva_nr99_v138.1_wSpecies_train_set.fa.gz", multithread = TRUE, minBoot = 80)

# Converting the taxonomy assignments to data frame so we can filter the taxonomic assignments.
o16S_taxa_original <- as.data.frame(o16S_taxa)

# Filtering out ASVs not assigned to a phyla
o16S_taxa_no_na <- o16S_taxa_original[-(which(is.na(o16S_taxa_original$Phylum))),]

# Filtering out ASVs not assigned to bacteria or archaea
o16S_taxa_bac <- o16S_taxa_no_na[grep("Bacteria|Archaea", o16S_taxa_no_na$Kingdom),]

# Filtering out ASVs assigned to a chloroplast
o16S_taxa_nochl <- o16S_taxa_bac[grep("Chloroplast", o16S_taxa_bac$Order, 
                                                invert = TRUE),]

# Filtering out ASVs assigned to a mitochondria
o16S_taxa_nohost <- o16S_taxa_nochl[grep("Mitochondria", o16S_taxa_nochl$Family, 
                                                   invert=TRUE),]

```

## **Taxonomy Merging and ASV Filtering**
The last step is to filter the ASV abundance table by merging the ASV table with the taxonomy assignments. This will remove ASVs from the count table that were not assigned to a fungal phyla or were identified as nematodes or plant sequences. 

```{r}
# Transposing the ASV table so that taxonomy can be added. 
t_o16S_seqtab_nochim <- t(o16S_seqtab_nochim)

# Merging the two tables together based on row name. 
t_o16S_seqtab_filt <- 
  t_o16S_seqtab_nochim[row.names(t_o16S_seqtab_nochim)%in%row.names(
    o16S_taxa_nohost),]

colnames(t_o16S_seqtab_filt) <- sub("16S_S[0-9].*", "16S", colnames(t_o16S_seqtab_nochim))

# Number of ASVs & number of sequences post-filtering
nrow(t_o16S_seqtab_filt)
sum(t_o16S_seqtab_filt)

# Merging taxonomy information into ASV table
t_o16S_seqtab_tax <- merge(t_o16S_seqtab_filt, 
                                o16S_taxa_nohost, by = "row.names")

#  Creating ASV labels and make these new row names.
asv_o16S <- as.character(c(1:nrow(t_o16S_seqtab_tax)))
asv_o16S <- paste0("OTU_16S_", labels(asv_o16S))

# Write with the sequences
write.table(t_o16S_seqtab_tax, "~/Desktop/R/incline/16S_soil_rhizo_rerun_labkey/results/o16S_otutab_sequences.csv", 
            row.names = TRUE, col.names = TRUE, sep = ',')

write.table(t_o16S_seqtab_tax, "~/Desktop/R/incline/16S_soil_rhizo_rerun_labkey/results/o16S_otutab_sequences.txt", 
            row.names = TRUE, col.names = TRUE, sep = '\t')

row.names(t_o16S_seqtab_tax) <- NULL
row.names(t_o16S_seqtab_tax) <- asv_o16S

write.table(t_o16S_seqtab_tax, "~/Desktop/R/incline/16S_soil_rhizo_rerun_labkey/results/o16S_otutab.csv", 
            row.names = TRUE, col.names = TRUE, sep = ',')

write.table(t_o16S_seqtab_tax, "~/Desktop/R/incline/16S_soil_rhizo_rerun_labkey/results/o16S_otutab.txt", 
            row.names = TRUE, col.names = TRUE, sep = '\t')

# couldn't figure out what this was for
# sort(colSums(select_if(t_o16S_rhizo_seqtab_tax, is.numeric)))

```
